---
title: "Sales performance analysis and economic drivers in retail"
author: "Jordy Alfaro Brenes"
date: "2025-03-13"
output: 
  html_document:
    toc: yes
    toc_float: yes
    df_print: paged
    code_folding: hide
---

# Introduction
In today's competitive retail industry, understanding what factors drive sales is essential for effective decision-making and strategic planning. This project aims to analyze historical sales data from one of the world’s largest retail chains to identify economic and seasonal variables that may influence business performance.

The dataset contains weekly sales records for multiple stores, along with additional information such as air temperature, fuel prices, holiday indicators, consumer price index (CPI), and unemployment rate. These variables allow us to explore patterns, assess external drivers of sales, and develop predictive models to forecast future performance under different economic scenarios.

By applying advanced data analysis techniques and machine learning models in R, this project seeks to provide actionable insights that can support inventory planning, marketing strategies, and resource allocation. The final results can help stakeholders anticipate changes in demand and make data-driven decisions that enhance operational efficiency and profitability.

# Dataset description

The dataset used in this project contains historical weekly sales data from a major retail chain in the United States. Each observation corresponds to a specific store and week, accompanied by several external and economic variables that may influence sales behavior.

The dataset includes the following variables:

- **Store**: Unique identifier for each store.
- **Date**: Start date of the sales week.
- **Weekly_Sales**: Total sales for the given store and week.
- **Holiday_Flag**: Indicates whether the sales week includes a national holiday (1 = holiday week, 0 = non-holiday).
- **Temperature**: Average regional air temperature (in Fahrenheit).
- **Fuel_Price**: Average fuel price in the region (in USD).
- **CPI**: Consumer Price Index, measuring changes in the price level of consumer goods and services.
- **Unemployment**: Regional unemployment rate (in percentage).

The dataset covers multiple years and stores, allowing for cross-sectional and time series analysis. It provides a rich context for exploring how economic and seasonal factors may influence retail performance. This makes it suitable for predictive modeling and scenario simulation in a business decision-making context.

```{r,include=FALSE}
library(ggplot2)
library(tidyverse)
library(lubridate)
library(corrplot)
library(GGally)
library(car)
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
library(xgboost)
library(knitr)
library(kableExtra)
```

```{r}
# 📁 Importar dataset
data <- read.csv("Walmart_Sales.csv")

# 👀 Vista rápida del dataset
head(data)
str(data)
summary(data)

# 🔄 Convertir la columna de fecha
data$Date <- dmy(data$Date)  # Usa lubridate para formato DD-MM-YYYY

# ➕ Crear nuevas variables: Año, Mes y Semana
data <- data %>%
  mutate(
    Year = year(Date),
    Month = month(Date, label = TRUE),
    Week = week(Date)
  )

data$Holiday_Flag <- as.factor(data$Holiday_Flag)
data$Store <- as.factor(data$Store)


# 💡 Verificar si hay valores NA
colSums(is.na(data))

# 📈 Tendencia general de ventas en el tiempo
ggplot(data, aes(x = Date, y = Weekly_Sales)) +
  geom_line(color = "steelblue") +
  labs(title = "Weekly Sales Over Time", x = "Date", y = "Weekly Sales")

# 📊 Boxplot: ventas en semanas con vs sin feriados
ggplot(data, aes(x = as.factor(Holiday_Flag), y = Weekly_Sales)) +
  geom_boxplot(fill = "darkorange") +
  labs(title = "Sales Distribution - Holiday vs Non-Holiday",
       x = "Holiday (0 = No, 1 = Yes)", y = "Weekly Sales")

# 🔍 Correlación entre variables numéricas
data_num <- data %>%
  select(Weekly_Sales, Temperature, Fuel_Price, CPI, Unemployment)

ggpairs(data_num)

# Histograma con curva de densidad
ggplot(data, aes(x = Weekly_Sales)) +
  geom_histogram(fill = "orange", bins = 30, color = "black") +
  geom_density(color = "darkorange", size = 1.2) +
  labs(title = "Distribution of Weekly Sales",
       x = "Weekly Sales", y = "Frequency") +
  theme_minimal()

# Cálculo de promedio por tienda
avg_sales <- data %>%
  group_by(Store) %>%
  summarise(Average_Weekly_Sales = mean(Weekly_Sales))

# Gráfico de barras
ggplot(avg_sales, aes(x = factor(Store), y = Average_Weekly_Sales)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Average Weekly Sales per Store",
       x = "Store", y = "Average Weekly Sales") +
  theme_minimal()

# Agrupar por semana
sales_by_date <- data %>%
  group_by(Date) %>%
  summarise(Total_Weekly_Sales = sum(Weekly_Sales))

# Gráfico de línea
ggplot(sales_by_date, aes(x = Date, y = Total_Weekly_Sales)) +
  geom_line(color = "orange", size = 1) +
  labs(title = "Total Weekly Sales Over Time",
       x = "Date", y = "Total Weekly Sales") +
  theme_minimal()

# Boxplot de ventas según Holiday_Flag
ggplot(data, aes(x = factor(Holiday_Flag), y = Weekly_Sales)) +
  geom_boxplot(fill = c("goldenrod", "orangered")) +
  labs(title = "Weekly Sales: Holiday vs Non-Holiday Weeks",
       x = "Holiday (0 = No, 1 = Yes)", y = "Weekly Sales") +
  theme_minimal()

# Gráficos individuales
ggplot(data, aes(x = Temperature, y = Weekly_Sales)) +
  geom_point(color = "darkorange") +
  labs(title = "Sales vs Temperature") + theme_minimal()

ggplot(data, aes(x = Fuel_Price, y = Weekly_Sales)) +
  geom_point(color = "darkorange") +
  labs(title = "Sales vs Fuel Price") + theme_minimal()

ggplot(data, aes(x = CPI, y = Weekly_Sales)) +
  geom_point(color = "darkorange") +
  labs(title = "Sales vs CPI") + theme_minimal()

ggplot(data, aes(x = Unemployment, y = Weekly_Sales)) +
  geom_point(color = "darkorange") +
  labs(title = "Sales vs Unemployment") + theme_minimal()

# Seleccionar solo variables numéricas
numeric_vars <- data %>%
  select(Weekly_Sales, Temperature, Fuel_Price, CPI, Unemployment)

# Calcular correlación
cor_matrix <- cor(numeric_vars)

# Heatmap con corrplot
corrplot(cor_matrix, method = "color", type = "upper", 
         addCoef.col = "black", tl.col = "black", number.cex = 0.8)

# Modelo de regresión lineal múltiple
modelo <- lm(Weekly_Sales ~ Temperature + Fuel_Price + CPI + Unemployment + Holiday_Flag, data = data)

# Resumen del modelo
summary(modelo)

# Instalar si no tienes el paquete
# install.packages("car")

vif(modelo)

# Graficar residuos y normalidad
par(mfrow = c(2, 2))
plot(modelo)
par(mfrow = c(1, 1))

# Predecir con el mismo dataset
data$Predicted_Sales <- predict(modelo, newdata = data)

# Comparar real vs predicho
ggplot(data, aes(x = Weekly_Sales, y = Predicted_Sales)) +
  geom_point(color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Actual vs Predicted Sales",
       x = "Actual Sales", y = "Predicted Sales") +
  theme_minimal()

# Dividir en train/test
set.seed(123)
index <- createDataPartition(data$Weekly_Sales, p = 0.8, list = FALSE)
train <- data[index, ]
test <- data[-index, ]

# Entrenar árbol
modelo_tree <- rpart(Weekly_Sales ~ Temperature + Fuel_Price + CPI + Unemployment + Holiday_Flag + Month + Store,
                     data = train, method = "anova")

# Graficar árbol
rpart.plot(modelo_tree)

# Predicción
pred_tree <- predict(modelo_tree, newdata = test)

# Evaluar
RMSE_tree <- RMSE(pred_tree, test$Weekly_Sales)
R2_tree <- R2(pred_tree, test$Weekly_Sales)

cat("Tree RMSE:", RMSE_tree, " | R²:", R2_tree, "\n")

modelo_rf <- randomForest(Weekly_Sales ~ Temperature + Fuel_Price + CPI + Unemployment + Holiday_Flag + Month + Store,
                          data = train, ntree = 500, importance = TRUE)

# Importancia de variables
varImpPlot(modelo_rf)

# Predicción
pred_rf <- predict(modelo_rf, newdata = test)

# Evaluación
RMSE_rf <- RMSE(pred_rf, test$Weekly_Sales)
R2_rf <- R2(pred_rf, test$Weekly_Sales)

cat("Random Forest RMSE:", RMSE_rf, " | R²:", R2_rf, "\n")


# Convertir data a matriz numérica
train_matrix <- model.matrix(Weekly_Sales ~ . -Date, data = train)[, -1]
test_matrix <- model.matrix(Weekly_Sales ~ . -Date, data = test)[, -1]

dtrain <- xgb.DMatrix(data = train_matrix, label = train$Weekly_Sales)
dtest  <- xgb.DMatrix(data = test_matrix, label = test$Weekly_Sales)

# Entrenar modelo
params <- list(objective = "reg:squarederror", eval_metric = "rmse")
modelo_xgb <- xgboost(params = params, data = dtrain, nrounds = 100, verbose = 0)

# Predicción
pred_xgb <- predict(modelo_xgb, newdata = dtest)

# Evaluación
RMSE_xgb <- RMSE(pred_xgb, test$Weekly_Sales)
R2_xgb <- R2(pred_xgb, test$Weekly_Sales)

cat("XGBoost RMSE:", RMSE_xgb, " | R²:", R2_xgb, "\n")


# Crear un dataframe resumen de métricas
model_performance <- tibble(
  Model = c("Linear Regression", "Decision Tree", "Random Forest", "XGBoost"),
  RMSE  = c(RMSE(predict(modelo, newdata = test), test$Weekly_Sales),
            RMSE_tree,
            RMSE_rf,
            RMSE_xgb),
  R2    = c(R2(predict(modelo, newdata = test), test$Weekly_Sales),
            R2_tree,
            R2_rf,
            R2_xgb)
)

# Gráfico comparativo - RMSE
ggplot(model_performance, aes(x = Model, y = RMSE, fill = Model)) +
  geom_bar(stat = "identity") +
  labs(title = "Model Comparison - RMSE", y = "RMSE", x = "") +
  theme_minimal() +
  theme(legend.position = "none")

# Gráfico comparativo - R²
ggplot(model_performance, aes(x = Model, y = R2, fill = Model)) +
  geom_bar(stat = "identity") +
  labs(title = "Model Comparison - R²", y = "R²", x = "") +
  theme_minimal() +
  theme(legend.position = "none")

# Tabla con kable para portafolio RMarkdown

model_performance %>%
  mutate(RMSE = round(RMSE, 2),
         R2 = round(R2, 4)) %>%
  kable(caption = "Model Performance Comparison", align = "c") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed"))

# Tomar una muestra real del test set como base
whatif_data <- test[1:10, ]

# Simular un aumento del precio del combustible
whatif_data$Fuel_Price <- whatif_data$Fuel_Price + 0.50

# Simular una disminución de temperatura
whatif_data$Temperature <- whatif_data$Temperature - 5

# Predicción bajo escenario modificado
pred_whatif_rf <- predict(modelo_rf, newdata = whatif_data)

# Comparar con predicción original
pred_original_rf <- predict(modelo_rf, newdata = test[1:10, ])

# Crear tabla comparativa
whatif_comparison <- tibble(
  Store = test$Store[1:10],
  Temperature_Original = test$Temperature[1:10],
  Fuel_Price_Original = test$Fuel_Price[1:10],
  Predicted_Original = round(pred_original_rf, 2),
  Temperature_Simulated = whatif_data$Temperature,
  Fuel_Price_Simulated = whatif_data$Fuel_Price,
  Predicted_Simulated = round(pred_whatif_rf, 2),
  Difference = round(pred_whatif_rf - pred_original_rf, 2)
)

print(whatif_comparison)

ggplot(whatif_comparison, aes(x = Store)) +
  geom_bar(stat = "identity", aes(y = Difference, fill = Difference > 0)) +
  labs(title = "Impact of Fuel Price ↑ and Temperature ↓ on Predicted Sales",
       y = "Change in Predicted Sales", x = "Store") +
  theme_minimal() +
  scale_fill_manual(values = c("red", "green"), labels = c("Decrease", "Increase"))


```